Supporting information: Tail associations and their impact on extinction risk

### 2019-11-05 MainText.Rmd
responses to Shyamolina's commnts

<!--concept for correlation in the extreme tail: tail association-->
Though earlier literature considered the influence of spatial synchrony 
in environmental variables on metapopulation extinction risk, to 
our knowledge the distinct influence of "asymmetric tail associations" between
environmental time series in different locations has hardly been considered. 
"Tail associations" are
associations between two random quantities in their extreme values. 
For instance, if two positively associated 
variables are more closely associated in their smaller values than in their larger values,
they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). 
If two positively associated variables are more associated in their larger 
values than in their smaller values, they are 
said to have stronger "right-tail" or "upper-tail" association 
(Figure \ref{fig_intro_cop}C, D). A precise definition of a measure of tail 
association is given in Methods.
Figure \ref{fig_intro_cop} shows that two variables can have a wide 
range of patterns of tail association, from extreme left- to extreme right-tail
association, with the same overall correlation. So tail association
is a distinct concept from correlation itself, and may have distinct ecological
ramifications [@Ghosh_copula]. Asymmetric tail associations have been documented 
between environmental variables, and between the same environmental variable 
measured in different places [@Serinaldi2008; @li2013; @She2018; @Goswami2018]. 
Asymmetric tail associations have also been documented between numerous ecological
variables and between environmental and ecological variables [@Ghosh_copula].

<!--ECEs probably produce tail associations-->
There are reasons to hypothesize that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, if a single ECE is associated
with spatially widespread, large values 
of some climatic variable, but small values of the
variable are associated with local weather rather than ECEs, the result could
be stronger right-tail association, through time, between measurements of the climatic variable
made in different locations.
For example, rainstorms which deposit exceptionally large amounts of rain
may do so over large areas; but periods with less rain may experience
scattered showers only, which may be much less correlated across locales.

@@

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Spearman's $\rho$ equal to $0.875$,
up to sampling variance), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\rho=0.875$ (up to sampling variance), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$, 
and we used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$, and we used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cumulative distribution functions of the desired beta and
gamma distributions were the transformations. Because these
were monotonic transformations, the Spearman correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix \ref{SM-makenoise}.

@@

We simulated our metapopulation models using environmental noises $\epsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variance), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variance), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\epsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cumulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Spearman correlation between $\epsilon_i(t)$ and $\epsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$; likewise partial Spearman correlations were the same. 
The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

@@

<!--Tuljapurkar approximation-->
Our results suggest that tail associations and asymmetries of tail associations
may have limited influence on the extinction risk of 
stage-structured, non-spatial models. Although we believe more research is 
certainly still needed
as to whether this conclusion holds more generally than the model we considered
(see below), if it does hold 
generally, it may be possible to obtain a better understanding of 
the result using a well-known approximation
of Tuljapurkar [@tulja1982; @tulja1990] for the long-term stochastic growth rate.
The long-term stochastic growth rate is an important quantity in demography
[@Caswell2000] and population viability analysis for conservation 
[@MorrisDoak2002], and it factors prominently into
extintion risk estimates for density-independent age- or 
stage-structured population models 
[@Lande1988], $\vec{P}(t+1)=M(t)\vec{P}(t)$. Here $\vec{P}(t)$ is the 
age- or stage-structured population and $M$ is a matrix-valued stationary stochastic 
process. The matrix entries of $M$ can correspond, for instance, to the fluctuating
fecundity and survival rates of the life stages of the population. 
For such a model, Tuljapurkar's approximation expresses the long-term stochastic 
growth rate as a function of the variances and covariances of the entries
of $M$ and of the eigenvalues and eigenvectors of the expected value of 
$M$. None of these quantities depends on tail associations between matrix entries.
Thus, whenever Tuljapurkar's approximation is a good
approximation, the long-term stochastic growth rate should not depend sensitively
on tail association. @Lande2003 provides reasoning supporting a claim that 
Tuljapurkar's approximation
is "likely to be accurate in many cases," (p. 60) though we are unaware of the extent
to which accuracy of the approximation has been tested for scenarios with
asymmetric tail associations.
We also note that this reasoning pertains only to density-independent models,
and says nothing about our results for the density-dependent version of our
stage-structured model.

>>>>

<!--concept for correlation in the extreme tail: tail association-->
Though earlier literature considered the influence of spatial synchrony 
in environmental variables on metapopulation extinction risk, to 
our knowledge the distinct influence of "asymmetric tail associations" between
environmental time series in different locations has hardly been considered. 
"Tail associations" are
associations between two random quantities in their extreme values. 
For instance, if two positively associated 
variables are more closely associated in their smaller values than in their larger values,
they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). 
If two positively associated variables are more associated in their larger 
values than in their smaller values, they are 
said to have stronger "right-tail" or "upper-tail" association 
(Figure \ref{fig_intro_cop}C, D). A precise definition of a measure of tail 
association is given in Methods.
<!--**DAN: You suggested we replace this with \nameref{Methods}. I do not care
one way or the other. You decide, we either do this for all such links or 
for none. Please decide yourself and then search through and make it that
way systematically across the ms. I point out though, links between pdfs will
not work.-->
Figure \ref{fig_intro_cop} shows that two variables can have a wide 
range of patterns of tail association, from extreme left- to extreme right-tail
association, with the same overall correlation. So tail association
is a distinct concept from correlation itself, and may have distinct ecological
ramifications [@Ghosh_copula]. Asymmetric tail associations have been documented 
between environmental variables, and between the same environmental variable 
measured in different places [@Serinaldi2008; @li2013; @She2018; @Goswami2018]. 
Asymmetric tail associations have also been documented between numerous ecological
variables and between environmental and ecological variables [@Ghosh_copula].

<!--ECEs probably produce tail associations-->
There are reasons to hypothesize that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, suppose a single ECE is associated
with spatially widespread, large values 
of some climatic variable. Also suppose that small values of the
variable are associated, instead, with local weather, which can differ from location
to location, rather than with ECEs. The result could
be stronger right-tail association, through time, between measurements of the climatic variable
made in different locations.
For example, rainstorms which deposit exceptionally large amounts of rain
may do so over large areas. But periods with less rain may experience
scattered showers only, which may be much less correlated across locales.

@@

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Spearman's $\rho$ equal to $0.875$,
up to sampling variance), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\rho=0.875$ (up to sampling variance), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$, 
and we used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$, and we used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cumulative distribution functions of the desired beta and
gamma distributions were the transformations. 
<!--***DAN: Shyamolina, you asked if I was talking about qbeta and qgamma here. 
Yes, but we do not need to say that - these are very standard functions in R. Anyone
who would be using R will know them, anyone not using R won't care that these
are the names of the functions in R. We just have to give what the mathematical
name is, which is "inverse 
cumulative distribution functions". Thanks for correcting my spelling of cumulative. 
Please delete this comment once you have read it.-->
Because these
were monotonic transformations, the Spearman correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix \ref{SM-makenoise}.

@@

We simulated our metapopulation models using environmental noises $\epsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variance), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variance), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\epsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cumulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. 
<!--***DAN: Shyamolina, you made a comment here in the pdf. I understand what you are saying.
What you did was mathematically equivalent to what I have written here, so it is OK we described it this way.-->
<!--***DAN: Shyamolina, in future please make comments in the Rmd, not attached to a pdf - that 
keeps everything in one place and makes it possible to reply to your comments more easily.-->
Because this was a monotonic transformation,
the Spearman correlation between $\epsilon_i(t)$ and $\epsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$; likewise partial Spearman correlations were the same. 
The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

@@

<!--Tuljapurkar approximation-->
Our results suggest that tail associations and asymmetries of tail associations
may have limited influence on the extinction risk of 
stage-structured, non-spatial models. Although we believe more research is 
certainly still needed
as to whether this conclusion holds more generally than the model we considered
(see below), if it does hold 
generally, it may be possible to obtain a better understanding of 
the result using a well-known approximation
of Tuljapurkar [@tulja1982; @tulja1990] for the long-term stochastic growth rate.
The long-term stochastic growth rate is an important quantity in demography
[@Caswell2000] and population viability analysis for conservation 
[@MorrisDoak2002], and it factors prominently into
extintion risk estimates for density-independent age- or 
stage-structured population models 
[@Lande1988], $\vec{P}(t+1)=M(t)\vec{P}(t)$. Here $\vec{P}(t)$ is the 
age- or stage-structured population and $M$ is a matrix-valued stationary stochastic 
process. The matrix entries of $M$ can correspond, for instance, to the fluctuating
fecundity and survival rates of the life stages of the population. 
For such a model, Tuljapurkar's approximation expresses the long-term stochastic 
growth rate as a function of the variances and covariances of the entries
of $M$ and of the eigenvalues and eigenvectors of the expected value of 
$M$. 
<!--***DAN: Shyamolina, I did not put it as a formula because that would have 
required a lot more text - we would have to define all the terms. We don't
actually need to know the specific dependence, just the fact that the growth
rate depends on the variances and covariances. -->
None of these quantities depends on tail associations between matrix entries.
Thus, whenever Tuljapurkar's approximation is a good
approximation, the long-term stochastic growth rate should not depend sensitively
on tail association. @Lande2003 provides reasoning supporting a claim that 
Tuljapurkar's approximation
is "likely to be accurate in many cases," (p. 60) though we are unaware of the extent
to which accuracy of the approximation has been tested for scenarios with
asymmetric tail associations.
We also note that this reasoning pertains only to density-independent models,
and says nothing about our results for the density-dependent version of our
stage-structured model.

### 2019-11-05 SuppMat.Rmd
responses to Shyamolina's commnts

<!--Spearman correlation of the extreme-tail-associated noise-->
The Spearman correlation of the extreme left- or right-tail associated 
noise was `r srho_common`, up to sampling variation.
<!--moderate tail association noise-->
Bivarite environmental noise with moderate left-tail association (as in Fig. \ref{MT-fig_intro_cop}B) was 
generated using a 2-dimensional Clayton copula. The statistical
theory of copulas has been introduced systematically in several texts 
(e.g., @nelsen2006_copula; @joe2014_dependence), and was briefly 
introduced by @Genest2007. Short, conceptual introductions aimed at ecological
audiences were provided by sections within the papers of @Anderson2018 and @Ghosh_copula. A
2-dimensional Clayton copula is equivalent to a particular bivariate 
probability density function with marginal distributions which are 
uniform on the interval $(0,1)$. The points on Fig. 
\ref{MT-fig_intro_cop}B were independent draws from the associated random variable.
The functions `claytonCopula` and `rCopula` from the `copula` package for the 
R programming language were used. The parameter of the Clayton family was
chosen so that the Spearman correlation was `r srho_common`, the same as that of the
extreme left-tail associated noise. This was accomplished with the `iRho`
function of the `copula` package. Noise with moderate right-tail
association was generated in a similar way but using the 2-dimensional 
survival Clayton copula.

>>>>

<!--Spearman correlation of the extreme-tail-associated noise-->
The Spearman correlation of the extreme left- or right-tail associated 
noise was `r srho_common`, up to sampling variation.
<!--***DAN: Dan to fill this in with a derivation of this fact-->

<!--moderate tail association noise-->
Bivarite environmental noise with moderate left-tail association (as in Fig. \ref{MT-fig_intro_cop}B) was 
generated using a 2-dimensional Clayton copula. The statistical
theory of copulas has been introduced systematically in several texts 
(e.g., @nelsen2006_copula; @joe2014_dependence), and was briefly 
introduced by @Genest2007. Short, conceptual introductions aimed at ecological
audiences were provided by sections within the papers of @Anderson2018 and @Ghosh_copula. A
2-dimensional Clayton copula is equivalent to a particular bivariate 
probability density function with marginal distributions which are 
uniform on the interval $(0,1)$. The points on Fig. 
\ref{MT-fig_intro_cop}B were independent draws from the associated random variable.
The functions `claytonCopula` and `rCopula` from the `copula` package for the 
R programming language were used. The parameter of the Clayton family was
chosen so that the Spearman correlation was `r srho_common`, the same as that of the
extreme left-tail associated noise. This was accomplished with the `iRho`
function of the `copula` package. Noise with moderate right-tail
association was generated in a similar way but using the 2-dimensional 
survival Clayton copula.
<!--***DAN: Shyamolina, in response to your comment in the pdf, what you
did is mathematically equivalent to this, so OK to say it this way. I added a 
bit of explanation of what that survival Clayon is. Please delete this
comment once you have read it.-->
The survival Clayton copula is a rotation of the Clayton copula by 180 degrees.
